---
title: "Quantium Chips Analysis"
author: "Priyanshi Singh"
date: "2025-06-19"
output: html_document
---


```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(here)
knitr::opts_chunk$set(echo = TRUE)

```

## Introduction

This analysis explores customer segments and chip purchasing behavior to support a strategic recommendation for the Category Manager. We aim to understand which segments purchase the most chips, how much they spend, and their product preferences.

## Load and Preview Data

```{r load-data}
# Load data
transactions <- read_csv(here("..", "data", "QVI_transaction_data.csv"))
customers <- read_csv(here("..", "data", "QVI_purchase_behaviour.csv"))

# Preview
glimpse(transactions)
glimpse(customers)

```

## Data Cleaning

```{r cleaning}
# Convert date
transactions$DATE <- as.Date(transactions$DATE)

# Remove outliers (optional threshold)
transactions <- transactions %>%
  filter(PROD_QTY > 0, PROD_QTY < 100)
  
# Check missing values
colSums(is.na(transactions))
colSums(is.na(customers))
```

## Feature Engineering (Brand & Pack Size)

```{r features}
# Brand is the first word in PROD_NAME
transactions <- transactions %>%
  mutate(BRAND = word(PROD_NAME, 1),
         PACK_SIZE = parse_number(PROD_NAME))

# Check top brands
transactions %>% count(BRAND, sort = TRUE)
```
## Merge with Customer Data

```{r merge}
data_combined <- merge(transactions, customers, by = "LYLTY_CARD_NBR")
glimpse(data_combined)
```

## Customer Segment Analysis

```{r segment-analysis}
segment_summary <- data_combined %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarise(
    total_sales = sum(TOT_SALES),
    avg_price = mean(TOT_SALES / PROD_QTY),
    transaction_count = n()
  ) %>%
  arrange(desc(total_sales))

segment_summary

segment_summary$LIFESTAGE <- factor(segment_summary$LIFESTAGE, levels = c(
  "YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES", "YOUNG FAMILIES", 
  "OLDER FAMILIES", "RETIREES", "NEW FAMILIES", "OLDER SINGLES/COUPLES"
))
```

## Visualization: Total Sales by Segment

```{r sales-plot}
ggplot(segment_summary, aes(x = LIFESTAGE, y = total_sales, fill = PREMIUM_CUSTOMER)) +
  geom_col(position = "dodge") +
  labs(title = "Total Chip Sales by Customer Segment", y = "Total Sales ($)", x = "Customer Lifestage") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90, hjust = 0.5))
```

##ï¸ Top Brands by Customer Segment

```{r top-brands}
top_brands <- data_combined %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER, BRAND) %>%
  summarise(total_sales = sum(TOT_SALES)) %>%
  arrange(LIFESTAGE, PREMIUM_CUSTOMER, desc(total_sales)) %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  slice_max(total_sales, n = 3)

top_brands
```

```{r top-brands-plot}
ggplot(top_brands, aes(x = reorder(BRAND, total_sales), y = total_sales, fill = BRAND)) +
  geom_col(show.legend = FALSE) +
  facet_grid(rows = vars(LIFESTAGE), cols = vars(PREMIUM_CUSTOMER), scales = "free") +
  coord_flip() +
  labs(title = "Top Brands per Segment", x = "Brand", y = "Sales") +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8)
  )

```

## ðŸ“¦ Pack Size Preferences

```{r pack-size}
pack_size_summary <- data_combined %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER, PACK_SIZE) %>%
  summarise(
    total_sales = sum(TOT_SALES),
    transactions = n()
  ) %>%
  arrange(desc(total_sales))

pack_size_summary %>% head(12)
```

```{r pack-size-plot}
ggplot(pack_size_summary, aes(x = PACK_SIZE, y = total_sales, color = PREMIUM_CUSTOMER)) +
  geom_line() +
  facet_wrap(~LIFESTAGE, scales = "free_y", ncol = 1) +  # Make one column vertically
  labs(title = "Sales by Pack Size", x = "Pack Size (g)", y = "Sales") +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10),
    
    
    axis.text = element_text(size = 8)
  )

```

## ðŸ’¼ Strategic Recommendations for Julia

Based on our insights:

- âœ… Focus on **Young Singles/Couples** and **Mainstream Midage** as they drive highest sales.
- âœ… Promote **larger pack sizes** and premium brands to segments that spend more per transaction.
- âœ… Use loyalty program data to promote **top 3 brands** per segment via email offers.
- âœ… Avoid low-volume brands for budget-conscious lifestages.

## ðŸ’¾ Export Summary Data (Optional)

```{r export}
write_csv(segment_summary, "output/segment_summary.csv")
write_csv(pack_size_summary, "output/pack_size_summary.csv")
write_csv(data_combined, "output/cleaned_data.csv")
```

## Conclusion

This report summarizes key insights on customer behavior and chip sales. It enables data-driven targeting of high-value segments, tailoring of product offerings, and strategic promotional planning.
